---
title: "Fuel Economy Dependence on Transmission Type"
author: "Matthew Gast"
date: "June 2015"
output:
  pdf_document: 
    fig_height: 3
  html_document:
    fig_caption: yes
    keep_md: yes
---

# Executive Summary

At first glance, manual transmissions seem to offer improved fuel
economy.  However, after controlling for the engine power and weight of the car, the effect of a manual transmission is not statistically significant.  The improved fuel economy of manual transmissions is apparently due to the increased prevalence of manual transmissions
in lighter cars with smaller engines.  The
message for car buyers is clear: if you want better fuel economy, purchase a lighter
car with a smaller engine, and pay only slight attention to the transmission
type.

# Exploratory Analysis

The data set analyzed in this paper, `mtcars`, consists of data on 32
automobiles across 10 dimensions. Begin by taking an exploratory look at the data, as shown by figures in the appendix.  The boxplot shows that in general, fuel economy is higher
for manual transmission cars, but of course, the purpose of this
analysis is to determine if that analysis is due to the transmission
type or other confounding factors.  For example, in the 1970s,
automatic transmissions were expensive extra-cost options on smaller
and therefore higher-mileage cars, but standard on larger low-mileage
cars.

```{r setup, echo=-(1:5), message=FALSE}
data(mtcars)
mtcars_raw <- mtcars
library(ggplot2)
library(gridExtra)
library(GGally)
```

# Regression Analysis

To assess the effect of transmission type, we will consider all the
available variables that can explain fuel economy.  The purpose of
this analysis is to determine whether the type of transmission affects
fuel economy.  Intuitively, the expectation is that a manual
transmission will increase fuel economy, and large cars and their
associated large engines will decrease fuel economy.

```{r cleanup, echo=FALSE}
mtcars$am <- as.factor(mtcars$am)
levels(mtcars$am) <- c("auto","manual")
mtcars$cyl <- as.factor(mtcars$cyl)
mtcars$vs <- as.factor(mtcars$vs)
levels(mtcars$vs) <- c("V","straight")
mtcars$gear <- as.factor(mtcars$gear)
mtcars$carb <- as.factor(mtcars$carb)
```

A pairwise plot of variables showing bivariate correlations is in the
appendix.  From the pairwise plot and calculated correlation values,
it is apparent that the number of cylinders, engine displacement, and power are all strongly
correlated.  In this analysis, we choose horsepower because that is the most commonly known statistic to buyers.

In addition to power, we expect that the weight of
the car matters to fuel economy because heavier cars are harder to
move.  Finally, it is common for mechanically simpler manual
transmissions to have more gears.  To control for whether the apparent
fuel economy of manual transmissions is due to having more forward
gears, we will include that in our modeling.

With potentially four predictors to choose from (transmission type,
power, weight, and the number of gears), we first
perform an analysis of variance between the four models.  The analysis of variance shows that adding the number of cylinders and
weight offer significant improvements to the model at the .01 level of
significance, but the number of gears does not.

```{r models}
fit1 <- lm (mpg ~ am, data=mtcars)
fit2 <- lm (mpg ~ am + hp, data=mtcars)
fit3 <- lm (mpg ~ am + hp + wt, data=mtcars)
fit4 <- lm (mpg ~ am + hp + wt + gear, data=mtcars)
anova(fit1, fit2, fit3, fit4)
```

Consider the results of the third model.  It explains 84% of the
variation in the data, and all coefficients except the transmission
type are significant at the .01 level.  The coefficients show that
increasing power decreases fuel economy, with each
10 horsepower resulting in a 3 mpg decrease.  Also as expected, increases
in weight decrease fuel economy, with each thousand pounds decreasing
fuel economy by about 2.9 mpg.  The coefficient on the transmission is
positive, well within a standard
deviation of zero, indicating that a manual transmission is expected to increase fuel economy by 2 mpg, but the effect is not statistically
signficant.

```{r model_summary}
summary(fit3)
```

Finally, we consider the robustness of the model.  In the appendix, a
residual plot shows that the residuals from the model are normally
distributed, with only one outlier on the residual plot.  However,
there are no strong single outliers that warp the model.  Not surprisingly, a high-performance Maserati is the biggest single influencer.

```{r}
head(sort(hatvalues(fit3),decreasing=TRUE),n=3)
```

\newpage

# Appendix of Figures

1.  Exploratory analysis - transmisssion type

This boxplot compares fuel economy by transmission type.  Manual
transmissions have a higher average, and the standard deviations of
manual and automatic transmissions do not overlap.  Manual
transmissions appear to have better fuel economy, but of course, this
plot does not take into account other factors.

```{r transmission_boxplot, echo=FALSE}
car_box <- ggplot(aes(x=am,y=mpg),data=mtcars)+geom_boxplot(aes(fill=am)) + xlab("Transmission Type") + ylab("Fuel Economy (mpg)") + ggtitle("Fuel Economy by Transmission") + scale_fill_discrete(name="Transmission\nType")
car_box
```

2.  Exploratory analysis - pairwise plots

The following pairwise plots show the bivariate correlation between
various factors.  When building a regression model, it is best to
avoid using predictors that are strongly correlated.  This pairwise
plot depicts correlation visually by plotting data in the lower left
side of the diagonal, as well as displaying the calculated correlation
in the upper right side of the diagonal.

These panel plots and correlation values can be used in selecting
predictors for regression models.  For example, there are strong
correlations between the number of cylinders, engine displacement, and
horsepower.  These correlations are not surprising, because larger
engines develop more power by using more cylinders.  When building a
regression model, it is best to avoid choosing highly correlated
predictors, however.

```{r pairwise_panel, messages=FALSE, fig.height=6, echo=FALSE}
pairpanel <- ggpairs(mtcars_raw,
   title="Bivariate Correlations in Automotive Descriptors",
   lower=list(continuous="smooth"),
   diag=list(continuous="blank"), 
   upper=list(params=list(corSize=4)),
   axisLabels="show")
pairpanel
```

3. Residual plot

```{r ggplot_qq, echo=FALSE, messages=FALSE}
gg_qq <- function(x, distribution = "norm", ... , label.outliers=FALSE, conf = 0.95 ){

  words <- strsplit(distribution, " ")[[1]]
  distName <- paste(toupper(substring(words, 1,1)),
  	            tolower(substring(words, 2)),
                    sep="", collapse=" ")

  q.function <- eval(parse(text = paste0("q", distribution)))
  d.function <- eval(parse(text = paste0("d", distribution)))
  x <- na.omit(x)
  ord <- order(x)
  n <- length(x)
  P <- ppoints(length(x))
  df <- data.frame(ord.x = x[ord], z = q.function(P, ...))

  Q.x <- quantile(df$ord.x, c(0.25, 0.75))
  Q.z <- q.function(c(0.25, 0.75), ...)
  b <- diff(Q.x)/diff(Q.z)
  coef <- c(Q.x[1] - b * Q.z[1], b)

  zz <- qnorm(1 - (1 - conf)/2)
  SE <- (coef[2]/d.function(df$z)) * sqrt(P * (1 - P)/n)
  fit.value <- coef[1] + coef[2] * df$z
  df$upper <- fit.value + zz * SE
  df$lower <- fit.value - zz * SE

  labels<-names(x)
  if(label.outliers==TRUE){ 
    df$label <- ifelse(df$ord.x > df$upper | df$ord.x < df$lower, labels[ord],"")
    }

  p <- ggplot(df, aes(x=z, y=ord.x)) +
    geom_point() + 
    geom_abline(intercept = coef[1], slope = coef[2]) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha=0.2) +
    ggtitle(paste(distName,"Q-Q Plot")) +
    xlab(paste("Theoretical",distName,"Quantile")) +
    ylab(paste("Sample",distName,"Quantile")) 
  if(label.outliers==TRUE) p <- p + geom_text( aes(label = label))
   return(p)
}
```

The residuals fitted to the model follow a normal distribution.  In
fact, there is only one outlier from a 95% confidence range, labeled
on the plot.^[The `gg_qq` function used here is customized from
*Foo0*'s post on StackOverflow:
http://stackoverflow.com/questions/4357031/qqnorm-and-qqline-in-ggplot2]

```{r resid_plot, echo=FALSE}
gg_qq(fit3$residuals,label.outliers=TRUE)
```